Topological Equivariant Artist Model

### 2021-03-09 draft/sections/next.tex
slides first draft

This work contributes a mathematical description of the mapping \vartist\ from data to visual representation. Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this mode; support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks. Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications for producing visualization where the structure and properties match those of the input data. These specifications are that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions. This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic. Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space.

>>>>

This work contributes a mathematical description of the mapping \vartist\ from data to visual representation. Combining Butler's proposal of a fiber bundle model of visualization data with Spivak's formalism of schema lets this model support a variety of datasets, including  discrete relational tables,, multivariate high resolution spatio temporal datasets, and complex networks. Decomposing the artist into encoding \vchannel, assembly \vmark, and reindexing \vindex\ provides the specifications for producing visualization where the structure and properties match those of the input data. These specifications are that the graphic must have continuity equivalent to the data, and that the visual characteristics of the graphics are equivariant to their corresponding components under monoid actions. This model defines these constraints on the transformation function such that they are not specific to any one type of encoding or visual characteristic. Encoding the graphic space as a fiber bundle provides a structure rich abstraction of the target graphical design in the target display space.

### 2021-03-09 slides/main.tex
slides first draft

\usepackage{tabularx}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{appendixnumberbeamer}

\usepackage{notation} % move this later

@@

\usepackage[backend=bibtex, style=numeric-comp, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
\bibliography{../draft/glasslab_viz.bib}

@@

\subsection{Background}
\begin{frame}{Tools are tuned to the continuity of the date \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}}
    \begin{columns}
        \column{0.33\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/data_formatting.png}
            \caption{Based on fig~2.5 in Munzner's VAD\cite{VisualizationAnalysisDesign}}
        \end{figure}

@@

\begin{frame}{Visualizations are (mostly) evaluated on equivariance}
    \begin{description}
        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
        \item[Effectiveness] design choices made in deference to perceptual saliency \cite (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
    \end{description}
\end{frame}

\begin{frame}{Models describe composition}
    \begin{description}
        \item[language model] APT: syntax and semantics of visualization (Mackinlay  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987})
        \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
        \item[category theory] the semiotics of visualization are commutative (Vickers \cite{vickersUnderstandingViz2013})
        \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
        \begin{columns}
            \column{.5\textwidth}

@@

\section{Prototype}

\section{References}
\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\appendix 
\begin{frame}{Rendering: Define a Pixel}
    \begin{columns}
        \column{0.5\textwidth}
        Given a pixel
        \begin{equation}
        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
        \end{equation}
        the inverse map of the bounding box 
        \begin{equation}
        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
        \end{equation}
        is a region $\gbase_p \subset \gbase$ such that 
        \begin{align}
            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
        \end{align}
        yields the color of the pixel. 
        \column{0.5\textwidth}
        \begin{figure}[H]
            \includegraphics[width=\textwidth]{figures/math/render.png}
        \end{figure}
    \end{columns}
\end{frame}

>>>>

\usepackage{tabularx}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{appendixnumberbeamer} 
\usepackage{minted}
\setminted[python]{fontsize=\scriptsize, 
                   linenos,
                   numbersep=8pt,
                   gobble=0,
                   frame=lines,
                   bgcolor=bg,
                   framesep=3mm} 
\usepackage{notation} % move this later

@@

\usepackage[backend=bibtex, style=numeric-comp, sorting=none, doi=false,isbn=false,url=false, giveninits=true]{biblatex}
\bibliography{../draft/glasslab_viz.bib}

@@

\subsection{Background}
\begin{frame}{Tools are tuned to the continuity of the date \cite{HeerSoftware2006,toryRethinkingVisualizationHighlevel2004}}
    \begin{columns}
        \column{0.33\textwidth}
        \begin{figure}
            \includegraphics[width=1\textwidth]{figures/intro/data_formatting.png}
            \caption{Based on fig~2.5 in Munzner's VAD\cite{munznerVisualizationAnalysisDesign2014}}
        \end{figure}

@@

\begin{frame}{Visualizations are (mostly) evaluated on equivariance}
    \begin{description}
        \item[Expressiveness] structure preserving mappings from data to graphic (Mackinlay \cite{mackinlayAutomatingDesignGraphical1986})
        \item[Effectiveness] design choices made in deference to perceptual saliency (Mackinlay \cite{clevelandResearchStatisticalGraphics1987,clevelandGraphicalPerceptionTheory1984,chambersGraphicalMethodsData1983a, munznerVisualizationAnalysisDesign2014})
        \item[Naturalness] easier to understand when properties match (Norman \cite{norman_things_smart})
        \item[Graphical Integrity] graphs show \textbf{only} the data (Tufte \cite{tufteVisualDisplayQuantitative2001})
    \end{description}
\end{frame}

\begin{frame}{Models describe composition}
    \begin{description}
        \item[language model] APT, GoG: syntax, semantics, and grammar of graphics (Mackinlay, Wilkenson  \cite{mackinlayAutomatingDesignGraphical1986, mackinlayAUTOMATICDESIGNGRAPHICAL1987,wilkinsonGrammarGraphics2005})
        \item[functional dependencies] constrained maps between data and visual representation(Sugibuchi \cite{sugibuchiFramwork2009}) 
        \item[category theory] the semiotics of visualization are commutative (Vickers \cite{vickersUnderstandingViz2013})
        \item[algebraic process] data ($\alpha$) and viz ($\omega$) transforms are symmetric (Kindlmann and Scheidegger \cite{kindlmannAlgebraicProcessVisualization2014})
        \begin{columns}
            \column{.5\textwidth}

@@

\section{Prototype}
\subsection{Artist}
\begin{frame}[fragile]{Artist}
    \begin{minted}{python}
        class ArtistClass(matplotlib.artist.Artist):
            def __init__(self, data, transforms, *args, **kwargs):
                # properties that are specific to the graphic
                self.data = data 
                self.transforms = transforms
                super().__init__(*args, **kwargs)

            def assemble(self, **args):
                # set the properties of the graphic

            def draw(self, renderer):
                # returns K, indexed on fiber then key 
                view = self.data.view(self.axes) 
                # visual channel encoding applied fiberwise 
                visual = {p: t['encoder'](view[t['name']])
                          for p, t in self.transforms.items()}
                self.assemble(**visual)
                # pass configurations off to the renderer
                super().draw(renderer)
        \end{minted}
\end{frame}

\begin{frame}[fragile]{Artists: Scatter \& Line}
\begin{figure}[H]
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/scatter_0.png}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
        \includegraphics[width=\textwidth]{figures/code/line_1.png}
    \end{subfigure}
\end{figure}
\begin{columns}
\column{,49\textwidth}
\begin{minted}{python}
fig, ax = plt.subplots()
artist = Point(data, transforms)
ax.add_artist(artist)
\end{minted}
\column{.49\textwidth}
\begin{minted}{python}
fig, ax = plt.subplots()
artist = Line(data, transforms)
ax.add_artist(artist)
\end{minted}
\end{columns}
\end{frame}

\begin{frame}[fragile]{Artists: Scatter \& Line}
\begin{minted}{python}
class Point(mcollections.Collection):
    def assemble(self, x, y, s, facecolors='C0' ):
        # construct geometries of the circle glyphs in visual coordinates
        self._paths = [mpath.Path.circle(center=(xi,yi), radius=si) 
                    for (xi, yi, si) in zip(x, y, s)] 
        # set attributes of glyphs, these are vectorized 
        # circles and facecolors are lists of the same size
        self.set_facecolors(facecolors)
\end{minted} 
    \pause
\begin{minted}{python}
class Line(mcollections.LineCollection):
    def assemble(self, x, y, color='C0'):
        #assemble line marks as set of segments 
        segments = [np.vstack((vx, vy)).T for vx, vy in zip(x, y)]
        self.set_segments(segments)
        self.set_color(color)
\end{minted}
\end{frame}

\begin{frame}[fragile]{Visual Transformations}
\begin{minted}{python}
cmap =  color.Categorical({'true':'deeppink', 'false':'deepskyblue'})
transforms = {'x': {'name': 'v4', 'encoder': lambda x: x},
                'y': {'name': 'v2', 'encoder': lambda x: x},
                'facecolors': {'name':'v3', 'encoder': cmap}, 
                's':{'name': None , 
                    'encoder': lambda _: itertools.repeat(.02)}}
\end{minted}
    \begin{itemize}
        \item \mintinline{python}{lambda x: x} is identity \vchannel\
        \item \mintinline{python}|{'name':None}| map into \vfiber\ without corresponding \dsection\
        \item \mintinline{python}{color.Categorical} is custom \vchannel
    \end{itemize}
\end{frame}

\begin{frame}[fragile]{Custom Complex \vchannel}
\begin{minted}{python}
class Categorical:
    def __init__(self, mapping):
        # check that the conversion is to valid colors
        assert(mcolors.is_color_like(color) for color in mapping.values())
        self._mapping = mapping

    def __call__(self, value):
        # convert value to a color
        return [mcolors.to_rgba(self._mapping[v]) for v in values]
\end{minted}
\pause
That we can test for action equivariance
\begin{minted}{python}
def test_nominal(values, encoder):
    m1 = list(zip(values, encoder(values)))
    random.shuffle(values)
    m2 = list(zip(values, encoder(values)))
    assert sorted(m1) == sorted(m2)
\end{minted}
\end{frame}
\subsection{Data}
\begin{frame}[fragile]{Fiber Bundle}
\begin{minted}{python}
@dataclass
class FiberBundle:
"""
Attributes
----------
K: {'tables': []}
F: {variable name: type}
"""
    K: dict 
    F: dict
\end{minted}
\end{frame}

\begin{frame}[fragile]{Discrete Connectivity}
\begin{minted}{python}
class VertexSimplex: #maybe change name to something else
    """Fiberbundle is consistent across all sections
    """
    FB = FiberBundle({'tables': ['vertex']},  
            {'v1': float, 'v2': str, 'v3': float})

    def __init__(self, sid = 45, size=1000, max_key=10**10):
        # create random list of keys
    def tau(self, k):
        # e1 is sampled from F1, e2 from F2, etc...
        return (k, (e1, e2, e3, e4))
\end{minted}
\end{frame}

\begin{frame}[fragile]{1D Continuous Connectivity}
\begin{minted}{python}
class EdgeSimplex: 
    FB = FiberBundle({'tables': ['vertex','edge']},
                         {'x' : float, 'y':  float, 
                         'color':mtypes.Color()}})
    def __init__(self, num_edges=4, num_samples=1000): 
        self.keys = range(num_edge) #edge id
        self.distances = np.linspace(0,1, num_samples)
        # half generlized representation of arcs on a circle
        self.angle_samples = np.linspace(0, 2*np.pi, len(self.keys)+1)
    @staticmethod
    def _color(edge):
        return ['red','orange', 'green','blue'][edge%4]
    @staticmethod
    def _xy(edge, distances, start=0, end=2*np.pi):
        # start and end are parameterizations b/c really there is 
        angles = (distances *(end-start)) + start
        return np.cos(angles), np.sin(angles)
    def tau(self, k): #will fix location on page on revision
        x, y = self._xy(k, self.distances, 
                        self.angle_samples[k], self.angle_samples[k+1]) 
        color = self._color(k) 
        return (k, (x, y, color))
\end{minted}
\end{frame}

\begin{frame}[fragile]{View}
\begin{minted}{python}
def view(self, axes):
    table = defaultdict(list)
    for k in self.keys:
        table['index'].append(k)
        for (name, value) in zip(self.FB.fiber.keys(), self.tau(k)[1]):
            table[name].append(value)
    return table
\end{minted}
\begin{description}
    \item [\mintinline{python}{VertexSimplex}] (name, value), value is scaler
    \item [\mintinline{python}{EdgeSimplex}]  (name, value), value is [x0, ..., xn]
\end{description}
\end{frame}
\begin{frame}[fragile]{Same Artist, Different Data Configurations}
    \begin{figure}[H]
        \begin{subfigure}{0.49\textwidth}
            \includegraphics[width=\textwidth]{figures/code/linec_1.png}
        \end{subfigure}
        \begin{subfigure}{0.49\textwidth}
            \includegraphics[width=\textwidth]{figures/code/lined_1.png}
        \end{subfigure}
    \end{figure}
\begin{minted}{Python}
simplex.GraphLine(FB, edge_table, vertex_table, 
                      connect=True)
simplex.GraphLine(FB, edge_table, vertex_table, 
                      num_samples=2, connect=False)
\end{minted}
\end{frame}

\section{Conclusion}
\begin{frame}{Summary}
    \begin{itemize}
        \item structure preserving maps from data to visual representation:
            \begin{itemize}
                \item data and graphics have equivalent continuity
                \item properties are equivariant under monoid actions
            \end{itemize}  
        \item fiber bundles with a schema are structure rich abstractions of 
            \begin{itemize}
                \item topologically complex heterogenous data 
                \item target display spaces
            \end{itemize}
        \item model can be iteratively integrated into existing Matplotlib architecture
    \end{itemize}
\end{frame}

\begin{frame}{Proposed Dissertation}
\begin{itemize}
    \item expansion of the mathematical framework to include worked out simple and complex addition
    \item formalization of definition of equivalance class \vartisteq
    \item implementation of artist with explicit \vindex\
    \item specification of interactive visualization
    \item mathematical formulation of a graphic with axes labeling
    \item implementation of new prototype artists that do not inherit from Matplotlib artists
    \item provisional mathematics and implementation of user level composite artists
    \item proof of concept domain specific user facing library 
\end{itemize}
\end{frame}

\section{References}
\begin{frame}[allowframebreaks]{References}
\printbibliography
\end{frame}
\appendix 
\begin{frame}{Rendering: Define a Pixel}
    \begin{columns}
        \column{0.5\textwidth}
        Given a pixel
        \begin{equation}
        p=\left[y_{top}, y_{bottom}, x_{right}, x_{left}\right]
        \end{equation}
        the inverse map of the bounding box 
        \begin{equation}
        \gbase_{p} ={\gsection_{xy}}^{-1}(p)
        \end{equation}
        is a region $\gbase_p \subset \gbase$ such that 
        \begin{align}
            \scriptstyle r_p &= \scriptstyle \iint\limits_{S_p} \rho_r(s)ds^{2}\\
            \scriptstyle g_p &= \scriptstyle \iint\limits_{S_p} \rho_g(s)ds^{2}\\
            \scriptstyle  b_p &= \scriptstyle \iint\limits_{S_p} \rho_b(s)ds^{2}
        \end{align}
        yields the color of the pixel. 
        \column{0.5\textwidth}
        \begin{figure}[H]
            \includegraphics[width=\textwidth]{figures/math/render.png}
        \end{figure}
    \end{columns}
\end{frame}
\begin{frame}[fragile]{GraphLine Data Model}
\begin{minted}{python}
class GraphLine:
    def __init__(self, FB, edge_table, vertex_table, num_samples=1000,
                        connect=False):
        #set args as attributes and generate distance
        if connect: # test connectivity if edges are continuous
            assert edge_table.keys() == self.FB.F.keys()
            assert is_continuous(vertex_table)

    def tau(self, k):
        # evaluates functions defined in edge table
        return(k, (self.edges[c][k](self.distances) 
                        for c in self.FB.F.keys()))

    def view(self, axes):
        # walk the edge_vertex table to return the edge function
        table = defaultdict(list)
        for (i, (start, end)) in sorted(zip(self.ids, self.vertices), 
                                            key=lambda v:v[1][0]):
            table['index'].append(i)
            # same as view for line, returns nested list
            for (name, value) in zip(self.FB.F.keys(), self.tau(i)[1]):
                table[name].append(value)
        return table
\end{minted}
\end{frame}
